Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof

### 2014-02-05 paper/Intuition.tex
finsih merging Sections. Few real changes, just bad auto merging. I may have lost some edits.

Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
follows. From the transitions of Figure \ref{trans}, a store may happen only
when a leaf cache is in \Mo{} state for that
address%(Invariant~\textit{processReq}).
Leaf caches cannot be ancestors of any
other node, so by Invariant \fullref{nonAncestorCompatible}, if a leaf cache is
in \Mo{} state, all other leaf caches will be in \In{} state preventing the
possibility of processing load and store requests for the same address at the
same time. Thus, Definition~\fullref{uniqStTimes} holds.

Once a response has been produced for a request $i$, the cache increments the
counter for the request it processes next. Thus there is at most one response
to each request, satisfying Definition~\fullref{uniqRespLabels}. As leaf caches
have no children, a leaf cache in at least \Sh{} state for address $a$ is
\clean{} for that address. Since processing a load request requires the cache
to be at least in the \Sh{} state, its response, which supplies the data for
that address present in the cache, supplies the latest value by
Invariant~\fullref{latestValue}.  Thus, the final condition for
Theorem~\fullref{storeAtomicity} is also satisfied. All that remains is to show
how these invariants hold.

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy,

Invariant \fullref{conservative} and the first part of Invariant
\fullref{localCompatibility} defines a $\ge$ relation between the states of a
parent and its child for any address.  Since the ancestor relation is the
reflexive-transitive closure of the parent relation, the $\ge$ relation is
preserved between the states of two caches $p$ and $c$ where $p$ is the
ancestor of $c$.

Any two caches $c_1, c_2$ neither of which are an ancestor of the other has a
lowest common ancestor $x$ which is an ancestor of both $c_1$ and $c_2$. $x$
has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
$c_1$ and $c_2$ respectively. For any address $a$, $x.\dstate[x_1][a]$ and
$x.\dstate[x_1][a]$ are compatible by the second part of
Invariant~\fullref{localCompatibility}. It follows that $x_1.\state[a]$ and
$x_2.\state[a]$ are compatible since these values are smaller than the
respective directory values.  From the previous paragraph about the relation
between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
are also compatible. This proves Invariant \fullref{nonAncestorCompatible}.

We prove Invariant \fullref{latestValue} by strong induction on time. Consider
two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
none of its children's directory entries can be in \Mo{}, all its descendants
should be no greater than \Sh{} for that address (because of Invariant
\fullref{conservative} and the fact that $\ge$ relation is preserved for the
state of a cache and that of its ancestor). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed in the previous
paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
compatibility means that non-descendants can not be in \Mo{} state. All of
$c$'s ancestors are non-leaves, so they can not process any store request.
Therefore, no cache save $c$ may have process a store request, which means that
$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
at time $t$.

If $c$ was not clean at time $t$, it must have received data from either its
parent or its child.  As can be seen in Figure~\ref{trans}, whenever data is
sent either from a parent to a child, or from a child to the parent, the sender
is \clean. By the induction hypothesis, it means that the sender has the latest
value for the address. Because of Invariant~\ref{noTransitWrite}, no store
request can be processed when the data is in transit, which means $c$ will have
the latest value for address $a$ at time $t+1$. This establishes Invariant
\fullref{latestValue}.

@@

If the value of the state of a child at the time of sending a response is the
same as the value of the directory for that child at the parent at the time of
receiving the response, it follows straightaway that the parent downgrades its
directory on receiving a response from the child. This is because a child
downgrades on sending a response and the next-states are synced on receiving
the response.  This proof uses exactly the same arguments presented above to
prove Invariant \fullref{conservative} -- we prove by considering the last
message sent or received at the child before $t$ (and the case when no message
has been sent or received). Similarly, we can show that a child upgrades its
state on receiving a response, satisfying the conditions in the previous paragraph.

Invariant \fullref{noTransitWrite} can be proved as follows. Consider a
response sent from a parent $p$ at time $t$ received by its child $c$ at time
$t'$. No other response can be received by the parent at time $t_p$ such that
$t+1 \le t_p \le t'$ in order to not violate Invariant \fullref{noCross}, so
$p.\dstate[c][a]$ at $t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can
only upgrade its directory for $c$. Similarly, no response can be sent by the
child at time $t_c$ such that $t \le t_c \le t'$ for the same reason and hence
$c.\state[a]$ at $t \le c.\state[a]$ at $t'$ since the child can only downgrade
its state.  This means that during $t_p$ such that $t+1 \le t_p \le t'$,
$p.\dstate[c][a] > \In$ and (hence $p.\state[a] > \In$ by the first part of
Invariant \fullref{localCompatibility}) since the parent upgraded its directory
for $c$ at $t$.  Similarly, during $t_c$ such that $t \le t_c \le t'$,
$c.\state[a] < \Mo$ since the child upgrades its state at $t'$. We have already
shown that (in the proof of Invariant \fullref{latestValue}) if any cache is
not in \In{} state, no non-descendent cache can process a store request. In
this case, during the times between $t$ and $t'$, $p$ is not in \In{} state.
Among $p$'s directory entries, all of them save the entry for $c$ is not in
\Mo{} state, so none of the descendents of $p$'s children, save $c$'s
descendents can process a store request. During the times between $t$ and $t'$,
$c$'s state is $< \Mo{}$, so neither $c$ nor its descendents can process a store
request, proving Invariant \fullref{noTransitWrite}.

>>>>

Given these invariants, we codify our intuition by showing that
Theorem~\fullref{storeAtomicity} follows naturally from these
invariants.  A store may happen only when leaf cache is in \Mo{} state
for that address (Invariant~\fullref{processReq}). Leaf caches cannot
be ancestors of any other node, so by Invariant
\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
other leaf caches will be in \In{} state preventing the possibility of
load and store requests for the same address (by Invariant
\fullref{processReq}), showing Definition~\fullref{uniqStTimes} holds.

Given Invariant~\fullref{noReqAgain}, once a response has been produced
for a request $i$, the cache may subsequently only process $j^{th}$
request such that $j > i$. Thus there is at most one response to each
request, satisfying Definition~\fullref{uniqRespLabels}. A load request
may be processed only if a cache is at least in \Sh{} state (Invariant
\fullref{processReq}). As leaf caches have no children, a leaf cache in
at least in \Sh{} state for address $a$ is \clean for that address. A
load response supplies the data for that address present in the cache,
which, by Invariant~\fullref{lastestValue}, contains the ``latest
value''. Thus, the final condition for Theorem~\fullref{storeAtomicity}
is also satisfied. All that remains is to show that these properties hold.

@@

For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
the cache tree hierarchy, from Invariants \fullref{conservative} and
\fullref{localCompatibility}, we can show that for any address $a$,
$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
relation is the reflective-transitive closure of the parent relation,
and Invariants \fullref{conservative} and the first part
\fullref{localCompatibility} of defines a $\ge$ relation between the
states of a parent and its child.

Any two caches $c_1, c_2$ neither of which are an ancestor of the
other will have a lowest common ancestor $x$ which is an ancestor of
both $c_1$ and $c_2$ and has two distinct children $x_1$ and $x_2$
such that they are the ancestors of $c_1$ and $c_2$ respectively. For
any address $a$, $x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are
compatible by the second part of
Invariant~\ref{localCompatibility}. From this, it straightaway follows
that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from
Invariant~\ref{conservative}). From the previous paragraph, it follows
that $c_1$ and $c_2$ are also compatible. This proves
Invariant \fullref{nonAncestorCompatibility}.

We prove Invariant \fullref{latestValue} by strong induction on
time. Consider two cases: cache $c$ is already \clean{} for an address
$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
value at time $t$. Since none of its children's directory entries can
be in \Mo{}, all its descendants should be no greater than \Sh{} for
that address (as discussed above). Any non-ancestor of $c$, and
non-descendant of $c$ must be compatible (as discussed above). Since
$c$ is \clean, it is at least in the \Sh{} state, so compatibility
means that non-descendants can not be in \Mo{} state. All of $c$'s
ancestors are non-leaves, so they can not process any store
request. Therefore, no cache save $c$ may have process a store
request, which means that $c$ contains the latest value for address
$a$ at time $t+1$ if it was \clean{} at time $t$.

If $c$ was not clean at time $t$, it must have received data from
either its parent or its child.  As can be seen in Figure~\ref{trans},
whenever data is sent either from a parent to a child, or from a child
to the parent, the sender is \clean. By the induction hypothesis, it
means that the sender has the latest value for the address. Because of
Invariant~\ref{noTransitWrite}, no store request can be processed when
the data is in transit, which means $c$ will have the latest value for
address $a$ at time $t+1$.

@@

If the value of the state of a child at the time of sending a response is the
same as the value of the directory for that child at the parent at the time of
receiving the response, it follows straightaway that the parent downgrades its
directory on receiving a response from the child, since a child downgrades on
sending a response and the next-states are synced on receiving the response.
The proof for this follows exactly the same arguments presented above to prove
Invariant \fullref{conservative} -- we prove by considering the last message
sent or received at the child before $t$ (and the case when no message has been
sent or received).

Invariant \fullref{noTransitWrite} can be proved as follows. Consider a response
sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
other response can be received by the parent at time $t_p$ such that $t+1 \le
t_p \le t'$ to not violate Invariant \fullref{noCross}, so $p.\dstate[c][a]$ at
$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
directory for $c$. Similarly, no response can be sent by the child at time
$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
child upgrades its state at $t'$. Using arguments similar to the proof of
Invariant \fullref{latestValue}, we can establish that no cache can be in \Mo{}
state for address $a$ when a response is in transit from $p$ to $c$.
Similarly, we can prove that no cache can be in \Mo{} state for address $a$
when a response is in transit for address $a$ from $c$ to $p$. Since data is in
transit at the same time as when the response is in transit, Invariant
\fullref{noTransitWrite} follows.
